**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**

### 2021-05-28 Manuscripts/Manuscript by chapters -v2/03-GAM Theory.Rmd
Updated files

To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure (\@ref(fig:l-q-response), C). The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time. The timeline can be divided in equally spaced _knots_, each knot being a region where a different basis function will be used. Because there are six timepoints for this group, five knots can be used. The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept). The choice of basis functions is already optimized in the package _mgcv_ depending on the number of knots. In Panel A of Figure \@ref(fig:basis-plot), the four basis functions (and the intercept) are shown. Each of the basis functions is composed of six different points (because there are six points on the timeline). To control the "wigliness" of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B. The penalty reduces the "wigliness" of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.

@@

#heatmap of the penalization coefficient
x_labels<-c("Intercept","1","2","3","4")
p13<-ggplot(basis_plot,
            aes(x=Basis,
                y=Basis))+
  geom_tile(aes(fill = cof), 
            colour = "black") +
    scale_fill_gradient(low = "white",
                        high = "#B50A2AFF")+ #color picked from KikiMedium
  labs(x='Basis',
       y='Basis')+
  scale_x_discrete(labels=x_labels)+
  geom_text(aes(label=round(cof,2)),
            size=7,
            show.legend = FALSE)+
  theme_classic()+
  theme(legend.title = element_blank())

@@

(ref:basis-plot-caption) Basis functions for a single smoother for time with five knots. A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown. B: Penalty matrix for the basis functions. Each basis function is penalized by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Penalized basis functions. Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis. D: Smoother for time and original data points. The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.

>>>>

To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure (\@ref(fig:l-q-response), C). The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time. The timeline can be divided in equally spaced _knots_, each knot being a region where a different basis function will be used. Because there are six timepoints for this group, five knots can be used. The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept). The choice of basis functions is already optimized in the package _mgcv_ depending on the number of knots. In Panel A of Figure \@ref(fig:basis-plot), the four basis functions (and the intercept) are shown. Each of the basis functions is composed of six different points (because there are six points on the timeline). To control the "wigliness" of the fit, each of the basis functions of Panel A is weighted by multiplying it by a coefficient according to the matrix of Panel B. The parameter estimates are penalized where the penalty reduces the "wigliness" of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.

@@

#heatmap of the  coefficients
x_labels<-c("Intercept","1","2","3","4")
p13<-ggplot(basis_plot,
            aes(x=Basis,
                y=Basis))+
  geom_tile(aes(fill = cof), 
            colour = "black") +
    scale_fill_gradient(low = "white",
                        high = "#B50A2AFF")+ #color picked from KikiMedium
  labs(x='Basis',
       y='Basis')+
  scale_x_discrete(labels=x_labels)+
  geom_text(aes(label=round(cof,2)),
            size=7,
            show.legend = FALSE)+
  theme_classic()+
  theme(legend.title = element_blank())

@@

(ref:basis-plot-caption) Basis functions for a single smoother for time with five knots. A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown. B: Matrix for basis function penalization. Each basis function is multiplied by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Weighted basis functions. Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis. D: Smoother for time and original data points. The smoother (line) is the result of the sum of each penalized basis function at each time point, with simulated values for the group shown as points.

### 2021-05-28 Manuscripts/Manuscript by chapters -v2/04-Longitudinal analysis with GAMs.Rmd
Making proposed changes in the manuscript
Updated files

The previous sections provided the basic theoretical framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to LMEMs or rm-ANOVA. This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using $\textsf{R}$ and the package `mgcv`. The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.

@@

The simulated data is based on the reported longitudinal changes in oxygen saturation ($\mbox{StO}_2$) in subcutaneous tumors that appear in Figure 3(C) in [@vishwanath2009]. In the paper, diffuse reflectance spectroscopy was used to quantify $\mbox{StO}_2$ changes in both groups at the same time points (days 0, 2, 5, 7 and 10). In the "Treatment" group (chemotherapy) an increase in $\mbox{StO}_2$ is observed through time, while a decrease is seen in the "Control" (saline) group. Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10% (matching the SD in the original paper). 
The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot), A and the inlet, respectively.

@@

An interaction model is typically the main interest in longitudinal biomedical data, as it takes into account the effect of treatment, time, and their combination. In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the _time_ effect to account for the repeated measures over time.  Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry. A detailed description on basis functions and correlations can be found in [@hefley2017].

For the data in Figure \@ref(fig:sim-smooth-plot),A it is of interest to know the effect that each treatment has in $\mbox{StO}_2$. The model then needs to incorporate independent smooths for _Group_ and _Day_, respectively. The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric ( _Day_ ) and the other is a factor ( _Group_ ). Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups. Using $\textsf{R}$ and the package `mgcv` the model syntax is:

`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5,bs="gp"), method='REML',data=dat_sim`

This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots. The smooth is constructed using gaussian process smooths, which is indicated by `bs="gp"`. These splines are used to model temporal trends and might be particularly suited for long-term studies where the correlation between measurements changes as a function of the time intervals [@simpson2018]. The parametric term _Group_ is added to quantify differences in the effect of treatment between groups, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017]. When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),B). Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_[@gratia]. A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].

@@

mult_plot<-f2+inset_element(
  f1, left = 0.01, 
  bottom = 0.5, 
  right = 0.5, 
  top = 1.0)+
  f3+f4+f6+
   plot_annotation(tag_levels='A')&
   ylim(c(-5,75)) &
  theme(
     text=element_text(size=18)
     )&
  thm

>>>>

The previous sections provided the basic framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs. This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using $\textsf{R}$ and the package `mgcv`. The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.

@@

The simulated data is based on the reported longitudinal changes in oxygen saturation ($\mbox{StO}_2$) in subcutaneous tumors that appear in Figure 3, C in [@vishwanath2009]. In the paper, diffuse reflectance spectroscopy was used to quantify $\mbox{StO}_2$ changes in both groups at the same time points (days 0, 2, 5, 7 and 10). In the "Treatment" group (chemotherapy) an increase in $\mbox{StO}_2$ is observed through time, while a decrease is seen in the "Control" (saline) group. Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10% (matching the SD in the original paper). 
The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot), A and the inlet, respectively.

@@

An interaction effect is typically the main interest in longitudinal biomedical data, as it takes into account treatment, time, and their combination. In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the _time_ effect to account for the repeated measures over time.  Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry. A detailed description on basis functions and correlations can be found in [@hefley2017].

For the data in Figure \@ref(fig:sim-smooth-plot), A the main effect of interest is how $\mbox{StO}_2$ changes over time for each treatment. To estimate this, the model needs to incorporate independent smooths for _Group_ and _Day_, respectively. The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric ( _Day_ ) and the other is a factor ( _Group_ ). Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups. Using $\textsf{R}$ and the package `mgcv` the model syntax is:

`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5,bs="gp"), method='REML',data=dat_sim)`

This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots. The smooth is constructed using gaussian process smooths, which is indicated by `bs="gp"`. These splines are used to model temporal trends and might be particularly suited for long-term studies where the correlation between measurements changes as a function of the time intervals [@simpson2018]. The parametric term _Group_ is added to quantify differences in the effect of treatment between groups, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017]. When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),B). Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_ [@gratia]. A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].

@@

mult_plot<-f2+inset_element(
  f1, left = 0.01, 
  bottom = 0.5, 
  right = 0.5, 
  top = 1.0)+
  f3+f4+f6+
   plot_annotation(tag_levels='A')&
   ylim(c(-7,75)) &
  theme(
     text=element_text(size=18)
     )&
  thm
